Simulations of gravity-driven flow of binary liquids in microchannels

### 2011-01-07 Text/paper.tex
Need only to change figures from desktop computer and submit it.

\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{shurik.kuzmin@gmail.com}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{DBR Schlumberger Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}

@@

After choosing the reference parameters, the grid refinement procedure
needs to keep the macroscopic parameters constant.  It is easy to check
that it yields the following quantity to
be independent of grid size
$H_{\mathrm{eff}}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
x } = \mathrm{const}$. We performed a number of simulations for the prescribed grids. The
simulation results in terms of grid numbers $N_x$, $N_y$, the film thickness $\delta$, center bubble
velocity $U_{bubble}$ are summarized in Table \ref{table:parameters:grid:refinement}.
The unified scaled profiles are shown in Fig. \ref{fig:grid:profiles}. One can
see that results converge for $H_{\mathrm{eff}}\geq 175$, and that with proper
initialization techniques, large enough time and different wall
wettabilities, results are different only in the 3rd digit even for
underresolved film thicknesses. To calculate how well the interface is
resolved, the ratio of the interface thickness to the film thickness is calculated. The
interface itself occupies approximately $5 \xi$, where
$\xi=\sqrt{k/A}=1$. The ratio of the interface thickness to the film thickness
$5\xi/H_{\mathrm{film}}$
 is shown in Table \ref{table:parameters:grid:refinement}. Based on these results one can conclude that
 the interface needs to be resolved as $40-50$ percent of the
expected film thickness for simulations to be grid independent. We further examine the
velocities in the center of the bubble to calculate the capillary number. One can see from Table
\ref{table:parameters:grid:refinement} that the bubble velocities are consistent and the
calculated capillary number corresponding to these velocities is $0.07$, which is slightly
larger than the capillary number we aimed for. The corresponding difference can be
attributed to the Poiseuille profile body force initialization.% and the
 %density and viscosity ratios, which are much larger in the
%Bretherton problem.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Grid/norm_grid_profs.eps}
\caption{Grid-refined profiles for the effective
channel widths
$H_{\mathrm{eff}}=100,125,150,175,200,225$.$\delta$ is scaled on $H_\mathrm{eff}$ and $\delta=0$
corresponds to the wall location. The profiles were taken at $x=14$ (nondimensional coordinates).
Capillary number is $\Ca=0.05$. The capillary number obtained from simulations $\Ca=0.07$. 
%One can seeconvergent results even for the underresolved film
%thickness.
\label{fig:grid:profiles}}
\end{figure}
%Here will be a consolidated table
\begin{table}
\begin{tabularx}{0.97\textwidth}{|X|X|X|X|X|X|}
\hline
$N_x$&$N_y$&$\delta$&$U_\mathrm{bubble}$&$\frac{5\xi}{H_{\mathrm{film}}}$&$N_\mathrm{iter}$\\
\hline
$1500$&$102$&$0.0694$&$0.0041$&$0.824$&$200000$\\
\hline
$1875$&$127$&$0.0688$&$0.0041$&$0.646$&$250000$\\
\hline
$2250$&$152$&$0.0676$&$0.0040$&$0.539$&$300000$\\
\hline
$2625$&$177$&$0.0679$&$0.0040$&$0.453$&$350000$\\
\hline
$3000$&$202$&$0.0668$&$0.0041$&$0.400$&$400000$\\
\hline
$3375$&$227$&$0.0663$&$0.0039$&$0.355$&$450000$\\
\hline
\end{tabularx}
\caption{The parameters and results for grid resolution. The simulated domain is
of size $N_x \times N_y$. $U_\mathrm{bubble}$ is the velocity of the bubble center measured at
$x=14$.  $5\xi$ is the interface thickness. $H_{\mathrm{film}}=\delta (N_y-2)$ is the size of the film in
lattice Boltzmann units. $Ca=0.05$ was taken to initialize the simulation.
\label{table:parameters:grid:refinement}}
\end{table}

@@

Next, we investigated the qualitative variation of the film thickness over the bubble length.
The comparison for the thickness variation is presented in Fig.~\ref{fig:sehgal:bubble:length}.
\begin{figure}
\includegraphics[width=0.47\textwidth]{Figures/Bubble/bubble_sehgal.eps}\hfill
\includegraphics[width=0.47\textwidth]{Figures/Bubble/bubble_length.eps}\\
\caption{The qualitative comparison for the film thickness across
the bubble length. Left (courtesy of \citet{sehgal-microchannel}),
right (present simulations); $x$ is scaled to $H_{\mathrm{eff}}$
and increases in the flow direction.\label{fig:sehgal:bubble:length}}
\end{figure}
One can see a qualitative agreement, i.e. the thickness increases towards the front meniscus and
rapidly decreases towards the rear meniscus. This shape is sometimes referred to as a "bullet" shape.

\subsection{The influence of different viscosities}
The results presented above are taken for the liquid viscosity ratio $10$.
We performed the same simulations for the liquid-gas viscosity ratio of $20$.
The relaxation parameters were taken as $\tau_{\mathrm{liquid}}=4.5$ and $\tau_{\mathrm{bubble}}=0.7$. We
did not find any significant differences from the results presented above. The
comparison for the liquid film thicknesses is presented in Fig.
\ref{fig:capillary:viscous}. The results validate our assumption that in
the low capillary flow regime the density ratio does not affect the film thickness,
and that the gas-liquid viscosity ratio of the order $10$ is sufficient to obtain results
consistent with the literature data.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Capillary_Viscous/capillaries_viscous.eps}
\caption{The film thicknesses versus capillary number for
$\frac{\mu_{\mathrm{liq}}}{\mu_{\mathrm{gas}}}=10$ and $\frac{\mu_{\mathrm{liq}}}{\mu_{\mathrm{gas}}}=20$. The
results show good consistency and mutual agreement.\label{fig:capillary:viscous}}
\end{figure}

\section{Conclusion}
This work presents numerical tips and a benchmark for the
Bretherton/Taylor problem using the binary liquid lattice Boltzmann method. The bubble was chosen sufficiently long for the film
thickness to
stabilize, and periodic boundary conditions were used to keep the simulations robust.  A bubble
train was simulated instead of the motion of a single bubble, and care was taken to minimize
the mutual influence of neighboring bubbles. The computational
results in terms of capillary number dependence and shape of the bubbles show consistency with the
previously published data. Surprisingly, with large enough viscosity ratio the results are
independent of any inertial effects.  An examination of the influence of grid resolution on the results allowed
us to determine that the phase interface should be resolved as at least $50$ percent of the simulated film thickness
in order for the simulations to be convergent. Though our results are specific to the binary liquid lattice
Boltzmann method, the numerical hints and procedures can be used for any
continuous interface method.
\bibliographystyle{model1-num-names}
\bibliography{paper}
%\bibliographystyle{unsrtnat}
%\bibliography{paper}
\end{document}

>>>>

\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{shurik.kuzmin@gmail.com}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{DBR Schlumberger Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}

@@

After choosing the reference parameters, the grid refinement procedure
needs to keep the macroscopic parameters constant.  It is easy to check
that it yields the following quantity to
be independent of grid size
$H_{\mathrm{eff}}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
x } = \mathrm{const}$. We performed a number of simulations for the prescribed grids. The
simulation results in terms of grid numbers $N_x$, $N_y$, the film thickness $\delta$, center bubble
velocity $U_{bubble}$ are summarized in Table \ref{table:parameters:grid:refinement}.
The unified scaled profiles are shown in Fig. \ref{fig:grid:profiles}. One can
see that results converge for $H_{\mathrm{eff}}\geq 175$, and that with proper
initialization techniques, large enough time and different wall
wettabilities, results are different only in the 3rd digit even for
underresolved film thicknesses. To calculate how well the interface is
resolved, the ratio of the interface thickness to the film thickness is calculated. The
interface itself occupies approximately $5 \xi$, where
$\xi=\sqrt{k/A}=1$. The ratio of the interface thickness to the film thickness
$5\xi/H_{\mathrm{film}}$
 is shown in Table \ref{table:parameters:grid:refinement}. Based on these results one can conclude that
 the interface needs to be resolved as $40-50$ percent of the
expected film thickness for simulations to be grid independent. We further examine the
velocities in the center of the bubble to calculate the capillary number. One can see from Table
\ref{table:parameters:grid:refinement} that the bubble velocities are consistent and the
calculated capillary number corresponding to these velocities is $0.07$, which is slightly
larger than the capillary number we aimed for. The corresponding difference can be
attributed to the Poiseuille profile body force initialization.% and the
 %density and viscosity ratios, which are much larger in the
%Bretherton problem.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Grid/norm_grid_profs.eps}
\caption{Grid-refined profiles for the effective
channel widths
$H_{\mathrm{eff}}=100,125,150,175,200,225$.$\delta$ is scaled on $H_{\mathrm{eff}}$ and $\delta=0$
corresponds to the wall location. The profiles were taken at $x=14$ (nondimensional coordinates).
Capillary number is $\Ca=0.05$. The capillary number obtained from simulations $\Ca=0.07$. 
%One can seeconvergent results even for the underresolved film
%thickness.
\label{fig:grid:profiles}}
\end{figure}
%Here will be a consolidated table
\begin{table}
\begin{tabularx}{0.97\textwidth}{|X|X|X|X|X|X|}
\hline
$N_x$&$N_y$&$\delta$&$U_\mathrm{bubble}$&$\frac{5\xi}{H_{\mathrm{film}}}$&$N_\mathrm{iter}$\\
\hline
$1500$&$102$&$0.0694$&$0.0041$&$0.824$&$200000$\\
\hline
$1875$&$127$&$0.0688$&$0.0041$&$0.646$&$250000$\\
\hline
$2250$&$152$&$0.0676$&$0.0040$&$0.539$&$300000$\\
\hline
$2625$&$177$&$0.0679$&$0.0040$&$0.453$&$350000$\\
\hline
$3000$&$202$&$0.0668$&$0.0041$&$0.400$&$400000$\\
\hline
$3375$&$227$&$0.0663$&$0.0039$&$0.355$&$450000$\\
\hline
\end{tabularx}
\caption{The parameters and results for grid resolution. The simulated domain is
of size $N_x \times N_y$. $U_\mathrm{bubble}$ is the velocity of the bubble center measured at
$x=14$.  $5\xi$ is the interface thickness. $H_{\mathrm{film}}=\delta (N_y-2)$ is the size of the film in
lattice Boltzmann units. $Ca=0.05$ was taken to initialize the simulation.
\label{table:parameters:grid:refinement}}
\end{table}

@@

Next, we investigated the variation of the film thickness over the bubble length.
The comparison for the thickness variation is presented in Fig.~\ref{fig:sehgal:bubble:length}.
\begin{figure}
\includegraphics[width=0.47\textwidth]{Figures/Bubble/bubble_sehgal.eps}\hfill
\includegraphics[width=0.47\textwidth]{Figures/Bubble/bubble_length.eps}\\
\caption{A qualitative comparison for the film thickness across
the bubble length. Left (courtesy of \citet{sehgal-microchannel}),
right (present simulations); $x$ is scaled to $H_{\mathrm{eff}}$
and increases in the flow direction.\label{fig:sehgal:bubble:length}}
\end{figure}
One can see a qualitative agreement, i.e. the thickness increases towards the front meniscus and
rapidly decreases towards the rear meniscus. This shape is sometimes referred to as a "bullet" shape.

\subsection{The influence of different viscosities}
The results presented above are taken for the liquid over gas kinematic viscosity ratio of $10$.
We performed the same simulations for the liquid over gas viscosity ratio of $20$.
The relaxation parameters were taken as $\tau_{\mathrm{liquid}}=4.5$ and $\tau_{\mathrm{bubble}}=0.7$. We
did not find any significant differences from the results presented above. The
comparison for the liquid film thicknesses is presented in Fig.
\ref{fig:capillary:viscous}. The results validate our assumption that in
the low capillary flow regime that the sufficiently high gas over liquid viscosity ratio ($\geq 10$) is sufficient to obtain results
consistent with the literature data.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Capillary_Viscous/capillaries_viscous.eps}
\caption{The film thicknesses versus capillary number for
$\frac{\mu_{\mathrm{liq}}}{\mu_{\mathrm{gas}}}=10$ and $\frac{\mu_{\mathrm{liq}}}{\mu_{\mathrm{gas}}}=20$. \label{fig:capillary:viscous}}
\end{figure}

\section{Conclusion}
This work presents numerical studies of the
Bretherton/Taylor problem using the binary liquid lattice Boltzmann method. The bubble was chosen sufficiently long for the film
thickness to
stabilize, and periodic boundary conditions were used to keep the simulations robust.  A bubble
train was simulated instead of the motion of a single bubble, and care was taken to minimize
the mutual influence of neighboring bubbles. The computational
results in terms of capillary number dependence and shape of the bubbles show consistency with the
previously published data. Surprisingly, with large enough viscosity ratio the results are
independent of any inertial effects.  An examination of the influence of grid resolution on the results allowed
us to determine that the phase interface should be resolved as at least $50$ percent of the film thickness
in order for the simulations to be grid independent. Though our results are specific to the binary liquid lattice
Boltzmann method, the numerical hints and procedures can be used for any
continuous interface method.
\section{Acknowledgement}
A.~Kuzmin wants to thank Schlumberger for their financial support.
\bibliographystyle{model1-num-names}
\bibliography{paper}
%\bibliographystyle{unsrtnat}
%\bibliography{paper}
\end{document}
